Derive final annotations based on NBAtlas reference

### 2025-08-18 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
WIP responding to reviews; not expected to render

@@

validation_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv"
```

@@

```{r}
validation_df <- readr::read_tsv(validation_url)

palette_df <- readr::read_tsv(palette_file)
celltype_pal <- palette_df$hex_color
names(celltype_pal) <- palette_df$NBAtlas_label

nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)
nbatlas_ontology_ids <- readr::read_tsv(nbatlas_ontology_file)
```

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation")
) |>
  tibble::rownames_to_column("cell_id") |>
  tidyr::separate(cell_id, into = c("library_id", "cell_barcode"), remove = FALSE) |>
  # keep UMAP for eventual viz
  dplyr::rename(
    consensus = consensus_celltype_annotation, 
    UMAP1 = UMAP.1, 
    UMAP2 = UMAP.2
  ) |>
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id")

@@

Define closely-related groups of cell types that will be used for resolving annotations:

```{r}
#### Vectors of closely-related NBAtlas immune cell types
nbatlas_t_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "T cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_nk_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "NK cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_myeloid_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Myeloid") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

# Note that cDC and Monocyte are not direct labels in the disseminated NBAtlas object,
# but these groupings are used in the paper's marker gene analysis.
nbatlas_cdc_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "cDC") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_monocyte_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Monocyte") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

#### Vectors of closely-related consensus cell types
consensus_t_celltypes <- validation_df |> 
  dplyr::filter(validation_group_annotation == "T cell") |>
  dplyr::pull(consensus_annotation)

consensus_endo_celltypes <- validation_df |> 
  dplyr::filter(validation_group_annotation == "endothelial cell") |>
  dplyr::pull(consensus_annotation)

consensus_myeloid_celltypes <- validation_df |> 
  dplyr::filter(
    # consider only groups corresponding to what NBAtlas breaks up 
    # the Myeloid group into in the tumor zoom
    validation_group_annotation %in% c("myeloid", "monocyte", "macrophage")
  )|>
  dplyr::pull(consensus_annotation)
```

@@

First, we'll label any cells that have exactly matching `scANVI` and `SingleR` labels.

```{r}
annotation_df <- celltype_df |>
  dplyr::select(cell_id, consensus, singler, scanvi) |>
  dplyr::mutate(
    cell_type_assignment = ifelse(
      scanvi == singler, scanvi, NA
    )
  )
```

@@

```{r}
sum(!is.na(annotation_df$cell_type_assignment)) / nrow(annotation_df)
```

@@

@@

>>>>

```{r}
# Source additional utilities functions:
# - subset_nbatlas_markers()
# - harmonize_celltypes()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

@@

# marker genes from NBAtlas
nbatlas_markers_file <- file.path(
  ref_dir, 
  "nbatlas-marker-genes.tsv"
)
```

@@

```{r}
palette_df <- readr::read_tsv(palette_file)
celltype_pal <- palette_df$hex_color
names(celltype_pal) <- palette_df$NBAtlas_label

label_map_df <- readr::read_tsv(nbatlas_label_map_file)
ontology_df <- readr::read_tsv(nbatlas_ontology_file)

nbatlas_markers_df <- readr::read_tsv(nbatlas_markers_file)
```

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation", "consensus_celltype_ontology")
) |>
  tibble::rownames_to_column("cell_id") |>
  tidyr::separate(cell_id, into = c("library_id", "cell_barcode"), remove = FALSE) |>
  # keep UMAP for eventual viz
  dplyr::rename(
    consensus = consensus_celltype_annotation, 
    consensus_ontology = consensus_celltype_ontology,
    UMAP1 = UMAP.1, 
    UMAP2 = UMAP.2
  ) |>
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id")

@@

@@

```{r}
ontology_slim_df <- ontology_df |>
  dplyr::select(-CL_annotation)

prep_for_annotation <- function(
    df, 
    annot_type, 
    ontology_df, 
    label_map_df) {

 df |>
    dplyr::select(all_of(c("cell_id", annot_type))) |>
    dplyr::rename(label = annot_type) |>
    # Join the family labels
    dplyr::left_join(label_map_df, by = c("label" = "NBAtlas_label_immune_zoom")) |>
    dplyr::rename(family_label = NBAtlas_label) |>
    # Ensure "Unknown" has a family label instead of NA for future-proofing
    dplyr::mutate(family_label = ifelse(is.na(family_label), "Unknown", family_label)) |>
    # Join in the ontology labels to match the **celltype** ontologies (label column)
    dplyr::left_join(ontology_slim_df, by = c("label" = "NBAtlas_label")) |>
    dplyr::rename(label_ontology = CL_ontology_id) |>
    # Join in the ontology labels to match the **family** ontologies (family_label column)
    dplyr::left_join(ontology_slim_df, by = c("family_label" = "NBAtlas_label")) |>
    dplyr::rename(family_ontology = CL_ontology_id) |>
    # We don't need the labels now that we have ontologies
    dplyr::select(-label, -family_label) |>
    # rename columns to start with `annot_type`
    dplyr::rename_with(\(x) {paste(annot_type, x, sep = "_")}, -cell_id)
}

singler_annotation_df <- prep_for_annotation(celltype_df, "singler", ontology_slim_df, label_map_df)
scanvi_annotation_df <- prep_for_annotation(celltype_df, "scanvi", ontology_slim_df, label_map_df)

annotation_df <- celltype_df |>
  dplyr::select(cell_id, consensus_ontology) |>
  dplyr::left_join(singler_annotation_df, by = "cell_id") |>
  dplyr::left_join(scanvi_annotation_df, by = "cell_id") |>
  dplyr::select(cell_id, consensus_ontology, starts_with("singler"), starts_with("scanvi"))
```

```{r}
annotation_df <- annotation_df |> 
  dplyr::mutate(
    CL_ontology_id = dplyr::case_when(
      # Use directly matching label
      scanvi_label_ontology == singler_label_ontology ~ scanvi_label_ontology, 
      # Check for family agreement between SingleR/scANVI
      scanvi_family_ontology == singler_family_ontology ~ scanvi_family_ontology,
      # Now use agreement with consensus to assign a label: first by label, and then by family
      singler_label_ontology == consensus_ontology ~ singler_label_ontology,
      scanvi_label_ontology == consensus_ontology ~ scanvi_label_ontology,
      singler_family_ontology == consensus_ontology ~ singler_family_ontology,
      scanvi_family_ontology == consensus_ontology ~ scanvi_family_ontology,
      # Everything else is NA
      .default = NA_character_
    )
  ) |>
  # keep only the final ontologies
  dplyr::select(cell_id, CL_ontology_id)

missing_cell_id <- annotation_df |>dplyr::filter(is.na(CL_ontology_id)) |>
  dplyr::pull(cell_id)

sjs_version <- readr::read_tsv(params$annotations_tsv)

stephyes_allyno <- sjs_version |>
  dplyr::filter(cell_id %in% missing_cell_id) |>
  dplyr::filter(!is.na(CL_ontology_id)) |>
  dplyr::pull(cell_id)

celltype_df |>
  dplyr::select(cell_id, singler, scanvi, consensus) |>
  dplyr::filter(cell_id %in% stephyes_allyno)
```

@@

```{r}
sum(!is.na(annotation_df$CL_ontology_id)) / nrow(annotation_df)
```

@@

#### Faceted by cell type assignment

This UMAP is the same as the previous one, except each facet highlights one specific cell type.

```{r fig.width = 16, fig.height = 16}
umap_df |>
  faceted_umap(
    annotation_column = cell_type_assignment,
    celltype_colors = celltype_pal
  )
```

@@

### Dot plot

This section shows dot plots of marker gene expression for the final annotations.
Specifically, we show the top 5 highest `log2FC` marker genes per NBAtlas cell type for validation.
Marker genes that are present in three or more cell types are not included.
Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.

```{r}
# data frame of top 5 upregulated genes per cell type
top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 5, "up") |>
  # only markers in no more than 2 categories
  # this is a good middle ground to ensure there are at least 2 markers per cell type
  dplyr::add_count(ensembl_gene_id) |>
  dplyr::filter(n < 3) |>
  dplyr::select(-n) 
```

```{r, fig.width = 24, fig.height = 12}

# get total number of cells per final annotation group and set up y_label
total_cells_df <- full_celltype_df |>
  dplyr::count(label, name = "total_cells") |>
  # order so cell types are grouped
  dplyr::inner_join(label_map_df, by = "label" ) |>
  dplyr::mutate(
    label_recoded = as.factor(label_recoded),
    label_recoded = forcats::fct_relevel(label_recoded, levels(label_recoded_factor))
  ) |>
  dplyr::arrange(label_recoded) |>
  dplyr::mutate(y_label = glue::glue("{label} ({total_cells})")) |>
  # rename as expected for dotplot function - label_recoded needs to be the main column
  dplyr::select(-label_recoded) |>
  dplyr::rename(label_recoded = label)

total_cells_df$y_label <- factor(total_cells_df$y_label, levels = rev(total_cells_df$y_label))
nbatlas_bar_order <- total_cells_df$label_recoded
```

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  # rename this after we've done all the wrangling
  singler_df = full_celltype_df |> dplyr::rename(label_recoded = label),
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = nbatlas_celltype_pal, 
  min_cells = 0
)
```
